Supporting information

### 2023-10-30 paper/main.tex

\author{Teddy Groves}
\affiliation{ DTU Biosustain,  }

\email{tedgro@biosustain.dtu.dk}
\author{Nicholas Luke Cowie}
\affiliation{ DTU Biosustain,  }

\author{Lars Keld Nielsen}
\affiliation{ DTU Biosustain,  }

@@

\title[]{Bayesian regression modelling facilitates quantitative
modelling of cell metabolism}
\makeatletter
\begin{document}
\maketitle
\begin{abstract}
This paper presents Maud, a command-line application that implements
Bayesian statistical inference for kinetic models of biochemical
metabolic reaction networks. Maud takes into account quantitative
information from omics experiments and background knowledge, as well as
structural information about kinetic mechanisms, regulatory interactions
and enzyme knockouts. Below, we review the existing options in this
area, explain how Maud improves on the state of the art, describe the
intended modelling workflow and illustrate its use with an example
application.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, breakable, enhanced, borderline west={3pt}{0pt}{shadecolor}, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

After specifying the qualitative aspects of the kinetic model, we
selected parameter values to use as ground truth by Monte Carlo sampling
using a previous model of the methionine cycle as a starting point (see
\citet{saa_general_2015} for this model).

We used these parameters to simulate steady states in a range of
plausible experimental conditions, again using \citet{saa_general_2015}
as a starting point. These steady states were then used to generate
simulated measurements using the measurement model described below in
Section~\ref{sec-methionine-measurement-model}.

@@

Running standard diagnostic checks indicated that the samples we
generated were from the target posterior distribution. The improved
\(\hat{R}\) statistic
\citep{vehtariRankNormalizationFoldingLocalization2021} for every
sampled variable was within 1\% of 1, indicating appropriate mixing
within and between Markov chains. Additionally, the number of effective
samples was high, indicating that we generated enough posterior samples
to support inferences about the bulks of the distributions of the
sampled parameters. Furthermore, we observed no post warm-up divergent
transitions, indicating that the sampler was able to transform the
log-posterior distribution, avoiding any regions with excessive
curvature that might inhibit exploration via HMC.

@@

>>>>

\author{Teddy Groves}
\affiliation{DTU Biosustain, DTU, Kongens Lyngby, Denmark}

\email{tedgro@biosustain.dtu.dk}
\author{Nicholas Luke Cowie}
\affiliation{DTU Biosustain, DTU, Kongens Lyngby, Denmark}

\author{Lars Keld Nielsen}
\affiliation{DTU Biosustain, DTU, Kongens Lyngby, Denmark}
\affiliation{   }

@@

\title[]{Bayesian regression modelling facilitates quantitative
modelling of cell metabolism}
\makeatletter
\begin{document}
\maketitle
\begin{abstract}
This paper presents Maud, a command-line application that implements
Bayesian statistical inference for kinetic models of biochemical
metabolic reaction networks. Maud takes into account quantitative
information from omics experiments and background knowledge, as well as
structural information about kinetic mechanisms, regulatory interactions
and enzyme knockouts. Below, we review the existing options in this
area, explain how Maud improves on the state of the art, describe the
intended modelling workflow and illustrate its use with an example
application.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, enhanced, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

After specifying the qualitative aspects of the kinetic model, we
selected parameter values to use as ground truth by Monte Carlo sampling
using a previous model of the methionine cycle as a starting point (see
\citet{saa_construction_2016} for this model).

We used these parameters to simulate steady states in a range of
plausible experimental conditions, again using
\citet{saa_construction_2016} as a starting point. These steady states
were then used to generate simulated measurements using the measurement
model described below in Section~\ref{sec-methionine-measurement-model}.

@@

Running standard diagnostic checks indicated that the samples we
generated were from the target posterior distribution. The improved
\(\hat{R}\) statistic
\citep{vehtariRankNormalizationFoldingLocalization2021} for every
variable of interest was within 2\% of 1, indicating appropriate mixing
within and between Markov chains. Additionally, the number of effective
samples was high, indicating that we generated enough posterior samples
to support inferences about the bulks of the distributions of the
sampled parameters. Furthermore, we observed no post warm-up divergent
transitions, indicating that the sampler was able to transform the
log-posterior distribution, avoiding any regions with excessive
curvature that might inhibit exploration via HMC.

@@

\hypertarget{supporting-information}{%
\section{Supporting information}\label{supporting-information}}

The document \texttt{supplementary.pdf} contains the following
information:

\begin{itemize}
\tightlist
\item
  Description of Maud's kinetic model including parameter dimensions and
  rate equations.
\item
  Description of the procedure used to generate the artificial case
  study datasets.
\item
  Specification of the prior distribution used for the case studies
\end{itemize}

### 2023-10-30 paper/supplementary.qmd

## Prior distributions compared with true parameter values

>>>>

Starting with the model in @saa_construction_2016, we extracted values for
enzyme concentrations, boundary conditions and fluxes. We used these values
to generate MCMC samples using Maud using the priors specified in section
@sec-methionine-priors. When this was finished, we selected one sample with
relatively high log probability to use as a ground truth in our case study.
These parameter values are shown below in table @tbl-case-study-params. We
manually inspected the parameter values to screen for any obviously implausible
values; we did not find any of these.

## Prior distributions compared with true parameter values {#sec-methionine-priors}
